\section{Overview}\label{sec:rules:overview}

Before doing a deep dive into the specific rules of funqual, let us look at an example.  Recall the save the pandas example from Section \ref{sec:background}.  It is reproduced in this Section as Figure \ref{lst:rules:pandasource} for convenience.  

\begin{figure}
    \begin{lstlisting}[language=C,gobble=8]
        int breed_and_release_pandas() {
            Panda *baby_panda = malloc(sizeof(Panda));
            release_panda(baby_panda);
        }

        int save_the_pandas() {
            stop_deforestation());
            if (pandas_are_saved()) {
                printf("Stopping deforestation saved the pandas!\n");
                return 1;
            }

            breed_and_release_pandas();
            if (pandas_are_saved()) {
                printf("Breeding pandas in captivation and releasing them has saved the pandas!\n");
                return 1;
            }
            return 0;
        }

        int main(void) {
            if (save_the_pandas()) {
                printf("The pandas have been saved!\n");
            }
        }
    \end{lstlisting}
    \caption{Example C program.  Running this code in a production environment may not actually save the pandas}
    \label{lst:rules:pandasource}
\end{figure}

Let us now imagine that there is some constraint whereby \lstinline{save_the_pandas} should not allocate memory. As programmers we would like to believe that we are disciplined enough to remember this rule and enforce it ourselves. In practice, self-regulation like this often ends poorly. As a result we would like a tool like funqual to enforce this constraint automatically.

\begin{sloppypar}
Funqual allows us as programmers to create our own type qualifiers and to apply whatever meaning we want to those qualifiers.  In this particular case we create two type qualifiers: \lstinline{static_memory} and \mbox{\lstinline{dynamic_memory}.} We also create one rule: $restrict\_indirect\_call(static\_memory, dynamic\_memory)$. When the programmer qualifies a function with \lstinline{static_memory}, that declares the intent that this function will \textit{never} allocate memory on the heap. When the programmer qualifies a function with \lstinline{dynamic_memory}, that declares the intent that this function always allocates memory on the heap\footnote{The meanings of these type qualifiers are as determined by the programmer; without a rule to operate on them, funqual will completely ignore the type qualifiers.}. The rule $restrict\_indirect\_call(static\_memory, dynamic\_memory)$ tells funqual that \lstinline{static_memory} functions are not allowed to call \lstinline{dynamic_memory} functions either directly or indirectly. If it is possible for a \lstinline{static_memory} function to reach a \lstinline{dynamic_memory} function, then the rule has been violated and funqual should inform the user.
\end{sloppypar}

In the example about saving the pandas, we would qualify \lstinline{save_the_pandas} as \lstinline{static_memory} and we would qualify \lstinline{malloc} as \lstinline{dynamic_memory}.  Figure \ref{fig:coloredpandacallgraph} shows the call graph for Figure \ref{lst:rules:pandasource} with \lstinline{static_memory} functions marked green and with \lstinline{dynamic_memory} functions marked red.

\input{figures/colored-panda-call-graph}

\begin{sloppypar}
By turning the program into a directed graph and by assigning types to the vertices, we have transformed the problem of type qualifier rule satisfaction into a graph problem.  A question like \textit{are there any static\_memory functions that inadvertently call dynamic\_memory functions} essentially boils down to \textit{are there any paths from green vertices to red vertices}.  In this example, the answer to that question is yes.  In the code, \lstinline{save_the_pandas} calls \lstinline{breed_and_release_pandas} which calls \lstinline{malloc} constituting an illicit call.  Equivalently, \lstinline{save_the_pandas} has an edge to \lstinline{breed_and_release_pandas} which has an edge to \lstinline{malloc} constituting an illicit path.  A well-typed program has no paths from green vertices to red vertices.  A poorly-typed program will have at least one path.  
\end{sloppypar}


